### 2023-10-29 tex/linear.tex
[ writing ] full draft
[ writing ] ready for submission

Languages and systems implementing features inspired by linear logic include
Rust~\citep{MK14,Rust}, ATS~\citep{Xi04,ZX05}, and various implementations of
session types~\citep{HLVCCDMPRTVTZ16}.
Additionally, forms of linearity have been used in theoretical work, for example
to bound computational complexity~\citep{GSS92,Hofmann03} or to write programs
which support incremental updates to computations~\citep{ER03,Ehrhard18}.
As for modality, \citet{IMO20} show that common data-flow analyses for
imperative languages can be recast as modal type systems, though such analyses
have generally been developed in an ad hoc fashion.
Such a recasting may be useful to understand the metatheoretic properties of
programs passing such static analyses.
More user-facing implementations of modal type systems include the system for
stack allocation in OCaml~\citep{DW22} and the approach to programmer-annotated
erasure found in Agda and Idris~\citep{Atkey18}.

In this chapter, I look at two standard type-theoretic features that motivate
considering calculi going beyond simple type theories as deliniated in
\cref{sec:simple} or equivalently by \citet{AACMM21}.
The first feature is a $\Box$-modality --- specifically, the $\Box$-modality of
intuitionistic S4, which I discuss in \cref{sec:modal}.
The second feature is linearity, which I discuss in \cref{sec:ill}.
In particular, I discuss possible syntaxes for each, with a mind to
being able to apply the techniques described in \cref{sec:simple}.
After having introduced these topics, I finish this chapter with a survey of
prior work on representing such type theories in proof assistants in
\cref{sec:linmech}.

>>>>

Languages and systems implementing features inspired by linear logic include
Rust~\citep{MK14,Rust}, Granule~\citep{Granule18}, ATS~\citep{Xi04,ZX05}, an
experimental Haskell extension~\citep{BBNPS17}, and various implementations of
session types~\citep{HLVCCDMPRTVTZ16}.
Additionally, forms of linearity have been used in theoretical work, for example
to bound computational complexity~\citep{GSS92,Hofmann03} or to write programs
which support incremental updates to computations~\citep{ER03,Ehrhard18}.
As for modality, \citet{IMO20} show that common data-flow analyses for
imperative languages can be recast as modal type systems, though such analyses
have generally been developed in an ad hoc fashion.
Such a recasting may be useful to understand the metatheoretic properties of
programs passing such static analyses.
More user-facing implementations of modal type systems include the system for
stack allocation in OCaml~\citep{DW22} and the approach to programmer-annotated
erasure found in Agda and Idris~\citep{Atkey18,Brady21}.

In this chapter, I look at two standard type-theoretic features that motivate
considering calculi going beyond simple type theories as delineated in
\cref{sec:simple} or equivalently by \citet{AACMM21}.
The first feature is a $\Box$-modality --- specifically, the $\Box$-modality of
intuitionistic S4, which I discuss in \cref{sec:modal}.
The second feature is linearity, which I discuss in \cref{sec:ill}.
In particular, I discuss possible syntaxes for each, with a mind to
being able to apply the techniques described in \cref{sec:simple}.
After having introduced these topics, I finish this chapter with a survey of
prior work on representing such type theories in proof assistants in
\cref{sec:linmech}.

### 2023-10-29 tex/linmech.tex
[ writing ] ready for submission

However, this adaptation of the DILL style does not obviously generalise to
semiring annotations.
Even for the multiplicative fragment, we seem to be working against the
direction of addition, instead using a subtraction operation whenever we use a
variable.
For exponentials, though, and particularly the $\oc$-introduction rule, what I
have done seems ad-hoc, not based on any pointwise algebraic operation.

@@

\subsection{Co-de-Bruijn syntax}

\Citet{McBride18} presents a mechanisation of the simply typed
$\lambda$-calculus in what he calls \emph{co-de-Bruijn} style.
He notes that syntax based on de Bruijn indices, as presented in
\cref{sec:simple}, finds a canonical way to place contractions and weakenings
by eagerly placing contractions wherever they could be needed (i.e., whenever a
rule has multiple premises) and leaving weakening as late as possible (i.e., at
the variable rule and at rules with no premises).
Co-de-Bruijn syntax, by contrast, finds a canonical way to place contractions
and weakenings by doing the reverse: weakening as early as possible (i.e., as
soon as the variable is bound) and contracting only where necessary.

Such a scheme straightforwardly adapts to multiplicative linear logic by
modifying the data structures presented by \citet{McBride18} to disallow all
contraction and weakening, as presented by \citet{RPKV20}.
With the additive rules, however, we get cases where variables appear multiple
times syntactically in a term but are still considered linear by the type
system, the simplest example being
$x : A \vdash \langle x, x\rangle : A \with A$.
Such rules are perhaps a new kind compared to what \citeauthor{McBride18}
considers, but it seems likely that just copying the same context to all the
premises would not be too hard to accommodate.
Meanwhile, we may consider implementing the $\oc$-modality as in DILL, with
intuitionistic variables handled using the regular co-de-Bruijn machinery from
the paper.
In summary, the co-de-Bruijn approach is promising for capturing linearity, but
has not been thoroughly investigated.

The other relevant contribution from \citet{RPKV20} is to recast the
context-splitting relations via connectives inspired by bunched
logic~\citet{oHP99}.
I adapt this method in \cref{sec:lnd}, but with a different notion of
context-splitting.

>>>>

However, this adaptation of the DILL style does not obviously generalise to
semiring annotations.
Even for the multiplicative fragment, we seem to be working against the
direction of addition, instead using a subtraction operation whenever we use a
variable.
An algebraic presentation of this fragment is given by \citet{ZD21}, in terms of
a partial functional cancellative addition operation.
For exponentials, though, and particularly the $\oc$-introduction rule, what I
have done seems ad-hoc, not based on any pointwise algebraic operation.

@@

\subsection{Co-De-Bruijn syntax}

\Citet{McBride18} presents a mechanisation of the simply typed
$\lambda$-calculus in what he calls \emph{co-De-Bruijn} style.
He notes that syntax based on De Bruijn indices, as presented in
\cref{sec:simple}, finds a canonical way to place contractions and weakenings
by eagerly placing contractions wherever they could be needed (i.e., whenever a
rule has multiple premises) and leaving weakening as late as possible (i.e., at
the variable rule and at rules with no premises).
Co-De-Bruijn syntax, by contrast, finds a canonical way to place contractions
and weakenings by doing the reverse: weakening as early as possible (i.e., as
soon as the variable is bound) and contracting only where necessary.

Such a scheme straightforwardly adapts to multiplicative linear logic by
modifying the data structures presented by \citet{McBride18} to disallow all
contraction and weakening, as presented by \citet{RPKV20}.
With the additive rules, however, we get cases where variables appear multiple
times syntactically in a term but are still considered linear by the type
system, the simplest example being
$x : A \vdash \langle x, x\rangle : A \with A$.
Such rules are perhaps a new kind compared to what \citeauthor{McBride18}
considers, but it seems likely that just copying the same context to all the
premises would not be too hard to accommodate.
Meanwhile, we may consider implementing the $\oc$-modality as in DILL, with
intuitionistic variables handled using the regular co-De-Bruijn machinery from
the paper.
In summary, the co-De-Bruijn approach is promising for capturing linearity, but
has not been thoroughly investigated.

The other relevant contribution from \citet{RPKV20} is to recast the
context-splitting relations via connectives inspired by bunched
logic~\citep{oHP99}.
I adapt this method in \cref{sec:lnd}, but with a different notion of
context-splitting.

### 2023-10-29 tex/mech-related.tex
[ writing ] ready for submission

\citet{Autosubst15} present the system \emph{Autosubst}, which provides various
tools for working with syntaxes with binding in the Coq proof assistant.
Autosubst is based on similar ideas to those \citeauthor{AACMM21} use:
de Bruijn-indexed terms with a distinguished variable rule and notion of
binding, acted upon by simultaneous renaming and substitution.

@@

In contrast, the work of \citet{HHLM22} is based around the category of contexts
under \emph{substitution} or, more precisely, the morphisms given by semantic
environments of whatever semantics is being defined.

\begin{itemize}
  \item Hard to bootstrap.
  \item All terms have the same, countably infinite context.
\end{itemize}

>>>>

\citet{Autosubst15} present the system \emph{Autosubst}, which provides various
tools for working with syntaxes with binding in the Coq proof assistant.
Autosubst is based on similar ideas to those \citeauthor{AACMM21} use:
De Bruijn-indexed terms with a distinguished variable rule and notion of
binding, acted upon by simultaneous renaming and substitution.

@@

In contrast, the work of \citet{HHLM22} is based around the category of contexts
under \emph{substitution} or, more precisely, the morphisms given by semantic
environments of whatever semantics is being defined.
The focus on substitution makes this work simpler than renaming-based
frameworks, because we get to avoid talking about renaming, whereas we always
ultimately need to talk about substitution.
However, what is a gain for internal simplicity is a loss for usability.
For the syntax, starting from substitution means that the framework provides no
help in \emph{proving} the admissibility of substitution, because the syntax is
not considered a model \emph{until} we can show that it admits substitution.
Similarly, for semantic models, we have to prove that each model we consider
respects semantic substitution, which is a stronger requirement than
respecting renaming.

Additionally, \citet{HHLM22} gain further simplicity of definitions by not
keeping track of contexts.
For example, untyped substitutions on a set (as opposed to scope-indexed family)
$X$ are functions from $\mathbb N$ to $X$.
Similarly, the simply typed families introduced in section 6 of the
aforementioned paper are indexed on their type, but not their context.
Ignoring scopes/contexts lets one talk about monads (as in
\emph{De Bruijn monads}) rather than a more complex notion like relative
monads~\citep{ACU15}.
However, it also means losing potentially useful and important information.

### 2023-10-29 tex/modal.tex
[ writing ] ready for submission

The methodology I use to produce a nicer syntax is that of \citet{judgmental}:
capturing the desired modality in the judgemental structure of the syntax.
To start, note the following two facts about a ``promotion'' rule similar to one
mentioned by \citet{BBdPH93}.

\begin{proposition}\label{thm:NKT4-P}
  In NJ + \TirName{N} + \TirName{K} + \TirName{4}, the following principle is
  admissible.
  \[
    \ebrule{%
      \hypo{\Box B_1, \ldots, \Box B_n \vdash A}
      \infer1[Promotion]{\Box B_1, \ldots, \Box B_n \vdash \Box A}
    }
  \]
\end{proposition}
\begin{proof}
  By induction on $n$.
  When $n = 0$, \TirName{Promotion} becomes just the necessitation rule \TirName{N}.
  For $n = \lvert\Gamma\rvert + 1$, we have the following derivation, where
  $\Box\Gamma$ abbreviates
  $\Box\Gamma_1, \ldots, \Box\Gamma_{\lvert\Gamma\rvert}$ and \TirName{4$'$} and
  \TirName{K$'$} abbreviate the obvious combinations of
  \TirName{4} and \TirName{K}, respectively, and \TirName{$\to$-E}.
  I silently apply weakening throughout the derivation for the sake of concision.
  \[
    \begin{prooftree}
      \hypo{\Box\Gamma, \Box B \vdash A}
      \infer1[$\to$-I]{\Box\Gamma \vdash \Box B \to A}
      \infer1[IH]{\Box\Gamma \vdash \Box\plr{\Box B \to A}}
      \infer0[Var]{\Box B \vdash \Box B}
      \infer1[4$'$]{\Box B \vdash \Box\Box B}
      \infer2[K$'$]{\Box\Gamma, \Box B \vdash \Box A}
    \end{prooftree}
  \]
\end{proof}

@@

The three new rules are listed in \cref{fig:PD}.
Let $\Gamma_v$ be the context $\Gamma$ but with all true hypotheses removed,
leaving only valid hypotheses.
Because I am working with a de Bruijn index-style presentation, weakening needs
to be admissible.
Therefore, the \TirName{$\Box$-I} rule needs to remove its true hypotheses,
rather than only applying in a context in which there are no true hypotheses.

>>>>

The methodology I use to produce a nicer syntax is that of \citet{judgmental}:
capturing the desired modality in the judgemental structure of the syntax.
To start, note the following two facts about a ``promotion'' rule similar to one
mentioned by \citet{BBdePH93}.

\begin{proposition}\label{thm:NKT4-P}
  When $\Gamma = B_1, \ldots, B_n$, let $\Box\Gamma$ abbreviate
  $\Box B_1, \ldots, \Box B_n$.
  Then, in NJ + \TirName{N} + \TirName{K} + \TirName{4}, the following principle
  is admissible.
  \[
    \ebrule{%
      \hypo{\Box\Gamma \vdash A}
      \infer1[Promotion]{\Box\Gamma \vdash \Box A}
    }
  \]
\end{proposition}
\begin{proof}
  By induction on $n$.
  When $n = 0$, \TirName{Promotion} becomes just the necessitation rule \TirName{N}.
  For $n = \lvert\Gamma\rvert + 1$, we have the following derivation, where
  \TirName{4$'$} and \TirName{K$'$} abbreviate the obvious combinations of
  \TirName{4} and \TirName{K}, respectively, with \TirName{$\to$-E}.
  I silently apply weakening throughout the derivation for the sake of concision.
  \[
    \begin{prooftree}
      \hypo{\Box\Gamma, \Box B \vdash A}
      \infer1[$\to$-I]{\Box\Gamma \vdash \Box B \to A}
      \infer1[IH]{\Box\Gamma \vdash \Box\plr{\Box B \to A}}
      \infer0[Var]{\Box B \vdash \Box B}
      \infer1[4$'$]{\Box B \vdash \Box\Box B}
      \infer2[K$'$]{\Box\Gamma, \Box B \vdash \Box A}
    \end{prooftree}
  \]
\end{proof}

@@

The three new rules are listed in \cref{fig:PD}.
Let $\Gamma_v$ be the context $\Gamma$ but with all true hypotheses removed,
leaving only valid hypotheses.
Because I am working with a De Bruijn index-style presentation, weakening needs
to be admissible.
Therefore, the \TirName{$\Box$-I} rule needs to remove its true hypotheses,
rather than only applying in a context in which there are no true hypotheses.

### 2023-10-29 tex/rensub.tex
[ writing ] ready for submission

Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
$\rho\,x = \sigma\,x$.
This follows from what we expect of equality of functions (function
extensionality).
Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
dividing up the usage annotations of $\Gamma$, and then a function producing
$\V$-values whose usage annotations come from that division.
Equality of $\Sigma$-types is tricky --- we need to equate the first components,
then rewrite the types of the second components by this equation before equating
them.
In practice, the equations rewriting other other equations build up so much that
I have given up on a first effort to give a treatment of the equational theory
of substitutions.
Note that recursively defined environments (\cref{def:lr-rec-env}) are also
$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
definition does not clearly help.

I hope that people working on substructural type systems in the future can take
inspiration from the process laid out in \cref{sec:lrkits} when working out the
appropriate notion of environment for their discipline.
Particularly, \cref{def:lr-rec-env} (the recursive definition) should serve as a
specification, if not the actual implementation, when coming across a new
substructural discipline.
As for the progression to \cref{def:lr-env}, this appears to arise from the fact
that the quantitative usage information is a refinement of the intuitionistic de
Bruijn index-based syntax.
%Being just a refinement means that the action of an environment on a variable
%is largely as it was in the intuitionistic case,

>>>>

Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
$\rho\,x = \sigma\,x$.
This follows from what we expect of equality of functions (function
extensionality).
Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
dividing up the usage annotations of $\Gamma$, and then a function producing
$\V$-values whose usage annotations come from that division.
Equality of $\Sigma$-types is tricky --- we need to equate the first components,
then rewrite the types of the second components by this equation before equating
them.
In practice, the equations rewriting other equations build up so much that
I have given up on a first effort to give a treatment of the equational theory
of substitutions.
Note that recursively defined environments (\cref{def:lr-rec-env}) are also
$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
definition does not clearly help.

I hope that people working on substructural type systems in the future can take
inspiration from the process laid out in \cref{sec:lrkits} when working out the
appropriate notion of environment for their discipline.
Particularly, \cref{def:lr-rec-env} (the recursive definition) should serve as a
specification, if not the actual implementation, when coming across a new
substructural discipline.
As for the progression to \cref{def:lr-env}, this appears to arise from the fact
that the quantitative usage information is a refinement of the intuitionistic De
Bruijn index-based syntax.
%Being just a refinement means that the action of an environment on a variable
%is largely as it was in the intuitionistic case,

### 2023-10-29 tex/semirings.tex
[ writing ] ready for submission

>>>>

This chapter and the following chapter re-present and expand the work of
\citet{WA21}.
For the thesis version, I have dropped mention of \emph{skew} semirings, which
allows the algebraic components to be more robust and better abstracted.
In particular, in \cref{sec:ren-sub-lr}, I talk about linear maps rather than
matrices, and define environments in terms of usage-checked variables rather
than raw well typed variables.

### 2023-10-29 tex/syntax.tex
[ writing ] ready for submission

Terms in this data type are difficult to write by hand, due to the
need for proofs that the usage contexts are handled correctly. For
example, the following term is needed to show that, in the $\{\gr0,
\gr1, \gr\omega\}$ (linearity) posemiring of \cref{def:lin-semiring},
$\oc\gr\omega$ forms a comonad.
Pattern synonyms \AgdaInductiveConstructor{$\multimap$I},
\AgdaInductiveConstructor{!E$'$}, and
\AgdaInductiveConstructor{!I$'$} stand for applications of
\AgdaInductiveConstructor{`con}, with the latter two taking explicit usage
contexts and proofs.
On concrete
posemirings (as in this example), unification is particularly poor at
inferring the usage contexts from the proofs because addition and
multiplication are no longer (definitionally) injective.
The function \AgdaFunction{var\#} is a way of turning a statically known de
Bruijn level and a usage proof into an application of \AgdaInductiveConstructor{`var}.

@@

%There is one more bunched combinator we have experimented with adding to the
%framework:
%
%\[
%  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
%  \times \plr{\grRprime \leq \gr0}
%  \times \plr{\grRprime \leq \grRprime + \grRprime}
%  \times T\,\grRprime
%\]
%
%The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
%of it, can be both discarded and duplicated indefinitely, and in the
%refinement we have a $T$.
%We use this combinator to introduce subterms that are used an unknown number of
%times, for example the continuations of the eliminator of an inductive type,
%or other fixed points.
%We can also use it in linear/non-linear style systems~\cite{Benton94} to make
%sure linear variables are not available in the intuitionistic fragment.
%
%Adding the $\Box$ combinator is the only thing we have found that requires our
%linear maps be functional rather than merely relational.

>>>>

Terms in this data type are difficult to write by hand, due to the
need for proofs that the usage contexts are handled correctly. For
example, the following term is needed to show that, in the $\{\gr0,
\gr1, \gr\omega\}$ (linearity) posemiring of \cref{def:lin-semiring},
$\oc\gr\omega$ forms a comonad.
Pattern synonyms \AgdaInductiveConstructor{$\multimap$I},
\AgdaInductiveConstructor{!E$'$}, and
\AgdaInductiveConstructor{!I$'$} stand for applications of
\AgdaInductiveConstructor{`con}, with the latter two taking explicit usage
contexts and proofs.
On concrete
posemirings (as in this example), unification is particularly poor at
inferring the usage contexts from the proofs because addition and
multiplication are no longer (definitionally) injective.
The function \AgdaFunction{var\#} is a way of turning a statically known De
Bruijn level and a usage proof into an application of \AgdaInductiveConstructor{`var}.
In the type, \AgdaPrimitive{$\infty$} is the ``infinite'' size, which all sizes
are less than.
Effectively, writing \AgdaPrimitive{$\infty$} allows us to ignore sizes when we
are not doing recursion involving sizes.

@@

%There is one more bunched combinator we have experimented with adding to the
%framework:
%
%\[
%  \plr{\Box T}\,\grR \coloneqq \Sigma\grRprime.~\plr{\grRprime \leq \grR}
%  \times \plr{\grRprime \leq \gr0}
%  \times \plr{\grRprime \leq \grRprime + \grRprime}
%  \times T\,\grRprime
%\]
%
%The idea of $\plr{\Box T}\,\grR$ is to assert that $\grR$, or some refinement
%of it, can be both discarded and duplicated indefinitely, and in the
%refinement we have a $T$.
%We use this combinator to introduce subterms that are used an unknown number of
%times, for example the continuations of the eliminator of an inductive type,
%or other fixed points.
%We can also use it in linear/non-linear style systems~\citep{Benton94} to make
%sure linear variables are not available in the intuitionistic fragment.
%
%Adding the $\Box$ combinator is the only thing we have found that requires our
%linear maps be functional rather than merely relational.

### 2023-10-29 tex/terms.tex
[ writing ] ready for submission

As part of the Curry-Howard correspondence, terms of a $\lambda$-calculus
correspond to derivations in a natural deduction system.
We could mechanise Gentzen's original definition of a natural deduction system
directly, but this definition is quite complicated.
In particular, if we want to give derivations an inductive definition, the use
of the discharge mechanism means that we actually need an inductive-inductive
type --- derivations, particularly those using $\to$-introduction, can involve
references to assumptions within their subderivations.
An inductive-inductive definition of derivations would complicate our programs
and proofs about natural deduction derivations, so I choose an alternative
representation.

Indeed, most authors since Gentzen, whether mechanising their work or not,
have opted to replace discharge of assumptions by explicit \emph{contexts} and
a variable rule.
Contexts can be justified as a way to keep track of undischarged assumptions.
In particular, we only produce derivations in the presence of a known collection
of \emph{free variables} specified by the context.
In other words, derivations are \emph{indexed} over their free variables and
their types.
When using an assumption within a derivation, we must say which free variable
it corresponds to.
Free variables are introduced by \emph{variable-binding} rules, like
$\to$-introduction.
\cref{fig:explicit-contexts} gives an example of the same derivation written
in Gentzen's style and in the explicit context style.

@@

Having chosen to use explicit contexts, the mechanisation must have a chosen
representation of contexts as a data structure.
While the notation in \cref{fig:explicit-contexts} uses names $f$ and $x$
for variables, I opt for a nameless representation.
In a nameless representation, variables are identified by their position in
the context, rather than by a name.
The absence of names means that $\alpha$-equivalence is just on-the-nose
equality, and also that we never have to reason about freshness of names.
Agda does not have support for nominal techniques~\cite{GP02}, which may have
made names a better option.

Most mechanisations choose contexts to be an inductive list of types.
However, I instead choose a functional, tree-shaped representation, as shown
with the type \AgdaRecord{Ctx}.
The type \AgdaDatatype{LTree} is the inductive type generated by leaves and
nullary \& binary nodes, and serves as a generalised ``length'' of the context.
The tree shape makes concatenation definitionally
injective, so that in cases where multiple new variables are bound in a subterm
(for example, $\otimes$-elimination), Agda's unification-based solving will
be more able to infer which variables have just been bound.
Within a given \AgdaBound{t}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}%
\AgdaDatatype{LTree}, we can define the positions of \AgdaBound{t} using
\AgdaDatatype{Ptr}.
A \emph{pointer} (\AgdaDatatype{Ptr}) into a tree picks out a leaf
(\AgdaInductiveConstructor{[-]}) following a path of lefts
(\AgdaInductiveConstructor{$\swarrow$}) and rights
(\AgdaInductiveConstructor{$\searrow$}) at any binary nodes encountered.

@@

The contents of the context --- the types --- are then stored in the functional
vector \AgdaField{ty-ctx}, which is a mapping from leaves in \AgdaField{shape}
to object language types \AgdaDatatype{Ty}.
The advantages of the functional vector representation will not become clear
until later chapters --- particularly the example in
\cref{sec:usage-elaborator}, where I make use of the ease of look-up and the
$\eta$-law of functions.
However, I claim for now that there is little to no disadvantage in the
functional vector representation --- in particular, we have no need for
function extensionality principles because we never talk about equality of
contexts.
For example, instead of using an equality of contexts to coerce a term, we can
use renaming.

\ExecuteMetaData[\Vectortex]{Vector}
\Ctx{}

>>>>

The main use I have for Agda in this thesis is to represent and reason about
programming languages.
In particular, I am interested in representing \emph{core languages} or
\emph{core calculi}.
Given a standard multi-pass compiler, the core language can be identified as the
first stage after which there can be no compile-time errors.
In other words, (some) terms of the core language arise from taking the source
code for a term, parsing it, scope-checking it, type-checking it, and doing
whatever other static checks are done to it.
Core language terms can also arise via optimisations on other core language
terms, or could not actually appear in the compilation of any real program, but
nonetheless be potentially handled by the optimiser and backend.
The core language can also be thought of as a representation of the
\emph{meaningful} part of a programming language, excluding (meaningless)
erroneous programs.

In most compilers, the core language is represented as an annotated abstract
syntax tree.
By construction, the abstract syntax tree representation makes syntactically
ill formed programs unrepresentable.
When representing a core language in a dependently typed language, including
Agda, we can take this idea further to make ill scoped, ill typed, and otherwise
meaningless programs unrepresentable.
Thus, we can truly \emph{define} a core language, and moreover reason
mathematically about it.

As is standard in the study of functional programming languages, I will take the
core languages I consider to be variants of the simply typed
$\lambda$-calculus~\citep{Church40,Barendregt93}.
As part of the Curry-Howard correspondence~\citep{Howard80}, terms of a typed
$\lambda$-calculus correspond to derivations in a natural deduction system, and
therefore we can take inspiration in terms of definitions and methodology from
logic when we come to mechanise core calculi in Agda.
We could mechanise Gentzen's original definition of a natural deduction system
directly, but this definition is quite complicated.
In particular, if we want to give derivations an inductive definition, the use
of the discharge mechanism means that we actually need an inductive-inductive
type --- derivations, particularly those using $\to$-introduction, can involve
references to assumptions within their subderivations.
An inductive-inductive definition of derivations would complicate our programs
and proofs about natural deduction derivations, so I choose an alternative
representation.

Indeed, most authors since Gentzen, whether mechanising their work or not,
have opted to replace discharge of assumptions by explicit \emph{contexts} and
a variable rule.
Contexts can be justified as a way to keep track of undischarged assumptions.
In particular, we only produce derivations in the presence of a known collection
of \emph{free variables} specified by the context.
In other words, derivations are \emph{indexed} over their free variables and
their types.
When using an assumption within a derivation, we must say which free variable
it corresponds to.
Free variables are introduced by \emph{variable-binding} rules, like
$\to$-introduction.
\Cref{fig:explicit-contexts} gives an example of the same derivation written
in Gentzen's style and in the explicit context style.

@@

Having chosen to use explicit contexts, the mechanisation must have a chosen
representation of contexts as a data structure.
While the notation in \cref{fig:explicit-contexts} uses names $f$ and $x$
for variables, I opt for a nameless representation.
In a nameless representation, variables are identified by their position in
the context, rather than by a name.
The absence of names means that $\alpha$-equivalence is just on-the-nose
equality, and also that we never have to reason about freshness of names.
Agda does not have support for nominal techniques~\citep{GP02}, which may have
made names a better option.

Most mechanisations choose contexts to be an inductive list of object-language
types.
However, I instead choose a functional, tree-shaped representation, as shown
with the type \AgdaRecord{Ctx}.
The type \AgdaDatatype{LTree} is the inductive type generated by
leaves (\AgdaInductiveConstructor{[-]}) and
nullary (\AgdaInductiveConstructor{$\upvarepsilon$}) \&
binary (\AgdaInductiveConstructor{\_<+>\_}) nodes, and serves as a generalised
``length'' of the context.
The tree shape makes concatenation definitionally
injective, so that in cases where multiple new variables are bound in a subterm
(for example, $\otimes$-elimination), Agda's unification-based solving will
be more able to infer which variables have just been bound.
Within a given \AgdaBound{t}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}%
\AgdaDatatype{LTree}, we can define the positions of \AgdaBound{t} using
\AgdaDatatype{Ptr}.
A \emph{pointer} (\AgdaDatatype{Ptr}) into a tree picks out a leaf
(\AgdaInductiveConstructor{[-]}) following a path of lefts
(\AgdaInductiveConstructor{$\swarrow$}) and rights
(\AgdaInductiveConstructor{$\searrow$}) at any binary nodes encountered.

@@

I use \AgdaDatatype{Ptr} to form a type family of \AgdaDatatype{LTree}-indexed
vectors \AgdaFunction{Vector}.
These vectors serve as the data structure containing the types, and in later
chapters usage information, of contexts.
The advantages of the functional vector representation will not become clear
until later chapters --- particularly the example in
\cref{sec:usage-elaborator}, where I make use of the ease of look-up and the
$\eta$-law of functions.
However, I claim for now that there is little to no disadvantage in the
functional vector representation --- in particular, we have no need for
function extensionality principles because we never talk about equality of
contexts.
For example, instead of using an equality of contexts to coerce a term, we can
use renaming.

\ExecuteMetaData[\Vectortex]{Vector}

The basic operations for building vectors from parts are \AgdaFunction{[\_]} to
create a singleton vector, \AgdaFunction{[]} to create an empty vector, and
\AgdaFunction{\_++\_} to append two vectors.

The shape of the context is usually not worth indexing over in term
representation, so I hide this index using the record type \AgdaRecord{Ctx}.
I also instantiate the element type of these vectors to \AgdaDatatype{Ty}, the
type of object-language types.

\Ctx{}

The three operators for building vectors lift to operations on \AgdaRecord{Ctx}
by suffixing \AgdaFunction{$^c$} --- giving us \AgdaFunction{[\_]$^c$},
\AgdaFunction{[]$^c$}, and \AgdaFunction{\_++$^c$\_}.
